\newlecture

\begin{instructor}
These notes are not complete.
\end{instructor}

Write the Scala wrapper for Java's array class

\begin{scalacode}
case class AnyArray[T](elts: Array[T])
\end{scalacode}

Methods get, set, length, and foreach

Write bitarray. These are the get and set methods

\begin{scalacode}
    def get(index: Int): Boolean = elts(index >> 5) >> (index & 0x1F) == 1

    def set(index: Int, value: Boolean) = {

      if (value) {
        elts(index >> 5) = elts(index >> 5) | (1 << (index & 0x1F))
      }
      else {
        elts(index >> 5) = elts(index >> 5) & ~(1 << (index & 0x1F))
      }
\end{scalacode}

Lesson: any code that uses BitArray needs to change. e.g., we can't write a generic foreach method

Introduce the ArrayLike trait:

\begin{scalacode}
  trait ArrayLike[T] extends Any {
    def get(index: Int): T
    def set(index: Int, value: T): Unit
    def length(): Int
  }
\end{scalacode}

Natural to write:

\begin{scalacode}
case class BitArray[T](elts: Array[Int]) ArrayLike[T]
\end{scalacode}

But, we can instead write:

\begin{scalacode}
  case class BitArray(elts: Array[Int]) ArrayLike[Boolean]
\end{scalacode}

  We've specialized the type parameter, but that's ok. (It's like what we did
  for \scalainline{List[Nothing]}.)

Further optimization: we are allocating an object that wraps the underlying
  Java array. Fix: use AnyVal:

\begin{scalacode}
  case class AnyArray[T](elts: Array[T]) extends AnyVal with CompactArray[T]
  case class BitArray(elts: Array[Int]) extends AnyVal with CompactArray[Boolean]
\end{scalacode}

Basically "sees through" the wrapper type.

Lessons: low-level optimizations are necessary for very high-performance code. Exposing low-level abstractions makes code un-reusable. GADTs can help.