
\newdiscussion{Implicits (Apr 6)}

This discussion will cover the first two questions of Homework 8, which asks you to use Scala implicits to wrap ``ugly'' Java interfaces with ``pretty'' Scala interfaces.

\section{What you need to do}

We will be completing HW 8, programming tasks 1 and 2 in class.  These two assignments are reproduced below.

\begin{enumerate}

  \item We should be able to construct paths by using the slash-operator to separate strings. For example, the expression \scalainline{"usr" / "bin" / "scala"} should be equivalent to the expression \scalainline{Paths.get("usr", "bin", "scala")}. 
  \item Similarly, given two paths, we should be able to join them using the slash-operator. For example, if \scalainline{p1} is \scalainline{"usr" / "local"} and \scalainline{p2} is \scalainline{"bin" / "scala"} then \scalainline{p1 / p2} should be equivalent to \scalainline{Paths.get("usr", "local", "bin", "scala")}. 
\end{enumerate}

\section{Hints}

First hint: this assignment requires \emph{neither} Scala implicit functions to work \emph{nor} Scala implicit function parameters.  You \emph{only} need Scala implicit classes.  If you do not remember the difference between these sorts of implicits, I strongly suggest that you familiarize yourself with their definitions.

Second hint: remember that Scala implicit classes require a few ingredients in order to work.  The most important ingredients are described here, but you can find the complete list of rules at: \url{https://www.artima.com/pins1ed/implicit-conversions-and-parameters.html#21.2}

\begin{enumerate}
  \item You must create a class and mark it as \scalainline{implicit}.
  \item Implicit classes must be defined inside an \scalainline{object}, \scalainline{trait}, or \scalainline{class}.
  \item The implicit class must be visible from the scope of the expression that you want rewritten.
  \item The compiler \emph{will not} attempt to rewrite an expression to use an implicit class if it type-checks as-is.  In other words, the expression must fail to type check \emph{unless} the compiler can find an implicit class in scope that causes the type check to succeed.
\end{enumerate}

\begin{instructor}
\section{Solutions}
  
Note that I have elided parts of the solution that are not directly related to the first two programming tasks.  The reason is that I want students to have to think about the rest of the solution on their own.

\begin{scalacode}
  import java.nio.file._

  object PathImplicits {  
    implicit class RichPath(p: Path) {
      def /(p2: String) : Path = p.resolve(p2)
      def /(p2: Path) : Path = p.resolve(p2)
    }

    implicit class RichString(s: String) {
      def /(s2: String) : Path = Paths.get(s).resolve(s2)
      def /(p2: Path) : Path = Paths.get(s).resolve(p2)
    }
  }
\end{scalacode}
  
\end{instructor}
